### 2012-10-13 Docs/Thesis/background.tex
Thesis: Background edits

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
as output and/or take functions as input. We say that functions are {\em first-class values} in a language.

Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.

FRP implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
and functions themselves ought to be reactive.

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.

@@

The earliest and still standard formulation of FRP is two primitive type
constructors: {\tt Behavior} and {\tt Event}, together with combinators that
produce values of this type. The easiest semantic definition for these types
is given in Figure~\ref{figure:classic_frp_semantic_types}

@@

\subsection{History of Classic FRP}
\label{subsection:Background-Classic_FRP-History_of_Classic_FRP}
Classic FRP was originally described as the basis of Fran~\cite{Elliott1997}
(Functional Reactive ANimation), a framework for declaratively specifying
interactive animations. Fran represented behaviors as two sampling functions,
one from a time to a value and a new behavior (so that history may be discared),
and the other from a time interval (a lower and upper time bound) to a value
interval and a new behavior. Events are represented as {\tt improving values},
which, when sampled with a time, produce a lower bound on the time to the next
occurrence, or the next occurrence if it has indeed occurred.

The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
of events and {\em bound properties} in the Beans framework, providing abstract
interfaces for FRP events and behaviors, and combinators as concrete classes
implementing these interfaces. The evaluation of Frapp\'{e} used Bean components
as sources and sinks, and the implementation of Bean events and bound properties
to propagate changes to the network~\cite{Courtney2001-2}.

@@

The Reactive~\cite{Elliott2009} system is a push-based FRP system with
first-class behaviors and events. The primary insight of Reactive is the
separation of reactivity (or changes in response to events whose occurrence time
could not be known beforehand) from time-dependence. This gives rise to {\em
reactive normal form}, which represents a behavior as a constant or simple
time-varying value, together with an event stream carrying values which are also
behaviors in reactive normal form. Push-based evaluation is achieved by forking
a Haskell thread to evaluate the head behavior, while waiting on the evaluation
of the event stream. Upon an event occurrence, the current behavior thread is
killed and a new thread spawned to evaluate the new behavior. Unfortunately, the
implmentation of Reactive uses a tenuous technique which depends on also forking
threads to evaluate two haskell values concurrently in order to implement event
merging. This relies on the library author to ensure consistency when this
technique is used, and leads to thread leakage when one of the merged events is
itself the merging of other events.

@@

An alternative approach to FRP was first proposed in work on
Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
{\em signal functions}. Arrows are abstract type constructors with input and
output type parameters, together with sequential ({\tt >>>}) and parallel
({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting
({\tt arr}) functions to the Arrow type. Signal functions are the first-class
abstraction in this approach, they represent time-dependent and reactive
transformers of events and signals,  which are themselves not first class,
since such values cannot be directly manipulated by the programmer. This
approach has two motivations: it increases modularity since both the input and
output of signal functions may be transformed, as opposed to signals or events
which may only be transformed in their output, and it avoids a large class of
time and space leaks which have emerged when implementing FRP with first-class
signals and events.

>>>>

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produce functions
as output and/or take functions as input. We say that functions are {\em first-class values} in a language.

Even in functional languages, reactive programs are generally written in an imperative style, using low-level and non-composable abstractions including callbacks
or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.

{\em Functional} Reactive Programming implies that a model should keep the characteristics of functional programming (i.e. that the basic constructs of the language
should remain first-class) while incorporating reactivity into the language model. In particular, functions should be lifted to operate on reactive values,
and functions themselves ought to be reactive.

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable. This has proven to be the main challenge in implementing FRP.

@@

The earliest and still standard formulation of FRP provides two primitive type
constructors: {\tt Behavior} and {\tt Event}, together with combinators that
produce values of these types. The easiest semantic definition for these types
is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list constructor {\tt []} should be considered to contain infinite as well as finite lists.}.

@@

\subsection{History of Classic FRP}
\label{subsection:Background-Classic_FRP-History_of_Classic_FRP}
Classic FRP was originally described as the basis of Fran~\cite{Elliott1997}
(Functional Reactive ANimation), a framework for declaratively specifying
interactive animations. Fran represented behaviors as two sampling functions,
one from a time to a value and a new behavior (so that history may be discared),
and the other from a time interval (a lower and upper time bound) to a value
interval and a new behavior. Events are represented as ``improving values'',
which, when sampled with a time, produce a lower bound on the time to the next
occurrence, or the next occurrence if it has indeed occurred.

The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
of events and ``bound properties'' in the Beans framework, providing abstract
interfaces for FRP events and behaviors, and combinators as concrete classes
implementing these interfaces. The evaluation of Frapp\'{e} used Bean components
as sources and sinks, and the implementation of Bean events and bound properties
to propagate changes to the network~\cite{Courtney2001-2}.

@@

The Reactive~\cite{Elliott2009} system is a push-pull FRP system with
first-class behaviors and events. The primary insight of Reactive is the
separation of reactivity (or changes in response to events whose occurrence time
could not be known beforehand) from time-dependence. This gives rise to {\em
reactive normal form}, which represents a behavior as a constant or simple
time-varying value, together with an event stream carrying values which are also
behaviors in reactive normal form. Push-based evaluation is achieved by forking
a Haskell thread to evaluate the head behavior, while waiting on the evaluation
of the event stream. Upon an event occurrence, the current behavior thread is
killed and a new thread spawned to evaluate the new behavior. Unfortunately, the
implmentation of Reactive uses a tenuous technique which depends on also forking
threads to evaluate two haskell values concurrently in order to implement event
merging. This relies on the library author to ensure consistency when this
technique is used, and leads to thread leakage when one of the merged events is
itself the merging of other events.

@@

An alternative approach to FRP was first proposed in work on
Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
{\em signal functions}. Arrows are abstract type constructors with input and
output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
Signal functions are the first-class
abstraction in this approach, they represent time-dependent and reactive
transformers of events and signals,  which are themselves not first class,
since such values cannot be directly manipulated by the programmer. This
approach has two motivations: it increases modularity since both the input and
output of signal functions may be transformed, as opposed to signals or events
which may only be transformed in their output, and it avoids a large class of
time and space leaks which have emerged when implementing FRP with first-class
signals and events.

\begin{figure}
\begin{code}
(>>>)  :: (Arrow a) => a b c -> a c d -> a b d
arr    :: (Arrow a) => (b -> c) -> a b c
first  :: (Arrow a) => a b c -> a (b, d) (c, d)
second :: (Arrow a) => a b c -> a (d, b) (d, c)
(***)  :: (Arrow a) => a b c -> a b d -> a b (c, d)
loop   :: (Arrow a) => a (b, d) (c, d) => a b c
\end{code}
\hrule
\caption{Arrow combinators.}
\label{figure:arrow_combinators}
\end{figure}
